---
title: " Information Dashboard Design: R methods"
format: html
author: "Erica Chan"
date: "26-03-2026"
date-modified: "last-modified"
number-sections: true
execute: 
  warnings: false
---

## Learning objectives:

create bullet chart by using ggplot2, create sparklines by using ggplot2 , build industry standard dashboard by using R Shiny.

## Getting started

### R Packages

The following code chunk imports the necessary R packages:

-   [tidyverse](https://www.tidyverse.org/) provides a collection of functions for performing data science task such as importing, tidying, wrangling data and visualising data. It is not a single package but a collection of modern R packages including but not limited to [readr](https://readr.tidyverse.org/), [tidyr](https://tidyr.tidyverse.org/), [dplyr](https://dplyr.tidyverse.org/), [ggplot](https://ggplot2.tidyverse.org/), [tibble](https://tibble.tidyverse.org/), [stringr](https://stringr.tidyverse.org/), [forcats](https://forcats.tidyverse.org/) and [purrr](https://purrr.tidyverse.org/).

-   [lubridate](https://r4va.netlify.app/chap31) provides functions to work with dates and times more efficiently.

-   [ggthemes](https://r4va.netlify.app/chap31) is an extension of ggplot2. It provides additional [themes](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/) beyond the basic themes of ggplot2.

-   [gtExtras](https://jthomasmock.github.io/gtExtras/index.html) provides some additional helper functions to assist in creating beautiful tables with [gt](https://gt.rstudio.com/), an R package specially designed for anyone to make wonderful-looking tables using the R programming language.

-   [reactable](https://glin.github.io/reactable/index.html) provides functions to create interactive data tables for R, based on the [React Table](https://github.com/tanstack/table/tree/v7) library and made with [reactR](https://github.com/react-R/reactR).

-   [reactablefmtr](https://kcuilla.github.io/reactablefmtr/index.html) provides various features to streamline and enhance the styling of interactive reactable tables with easy-to-use and highly-customizable functions and themes.

```{r}
pacman::p_load(lubridate, ggthemes, reactable,
               reactablefmtr, gt, gtExtras, tidyverse)
```

### Data

A personal database in Microsoft Access mdb format will be used as the data for this exercise.

`odbcConnectAccess()` of [**RODBC**](https://cran.r-project.org/web/packages/RODBC/index.html) package is used below to import the database query table into R.

```{R}
library(RODBC)
con <- odbcConnectAccess2007('data/Coffee Chain.mdb')
coffeechain <- sqlFetch(con, 'CoffeeChain Query')
write_rds(coffeechain, "data/CoffeeChain.rds")
odbcClose(con)
```

Next we will aggregate Sales and Budgeted Sales at the product level using the below code chunk.

```{r}
product <- coffeechain %>%
  group_by(`Product`) %>%
  summarise(`target` = sum(`Budget Sales`),
            `current` = sum(`Sales`)) %>%
  ungroup()
```

## Bullet chart in ggplot2

The code chunk below is used to plot bullet charts using ggplot2 functions. A bullet graph is a type of chart represented by a bar marked with extra encodings to show progress towards a goal or performance against a reference line.

```{r}
ggplot(product, aes(Product, current)) + 
  geom_col(aes(Product, max(target) * 1.01),
           fill="grey85", width=0.85) +
  geom_col(aes(Product, target * 0.75),
           fill="grey60", width=0.85) +
  geom_col(aes(Product, target * 0.5),
           fill="grey50", width=0.85) +
  geom_col(aes(Product, current), 
           width=0.35,
           fill = "black") + 
  geom_errorbar(aes(y = target,
                    x = Product, 
                    ymin = target,
                    ymax= target), 
                width = .4,
                colour = "red",
                size = 1) +
  coord_flip()
```

## Sparklines in ggplot2

### Data Preparation

The following code chunk summarises total sales by month from 2013 onward

```{r}
sales_report <- coffeechain %>%
  filter(Date >= "2013-01-01") %>%
  mutate(Month = month(Date)) %>%
  group_by(Month, Product) %>%
  summarise(Sales = sum(Sales)) %>%
  ungroup() %>%
  select(Month, Product, Sales)
```

The following code chunk computes the min, max and end of the month sales

```{r}
mins <- group_by(sales_report, Product) %>% 
  slice(which.min(Sales))

maxs <- group_by(sales_report, Product) %>% 
  slice(which.max(Sales))

ends <- group_by(sales_report, Product) %>% 
  filter(Month == max(Month))
```

The following code chunk computes the 25 and 75 percentile values

```{r}
quarts <- sales_report %>%
  group_by(Product) %>%
  summarise(quart1 = quantile(Sales, 
                              0.25),
            quart2 = quantile(Sales, 
                              0.75)) %>%
  right_join(sales_report)
```

### creating sparklines visualisation

the code below creates the sparkline graph using ggplot2 functions

```{r}
#| fig-width: 8
ggplot(sales_report, aes(x=Month, y=Sales)) + 
  facet_grid(Product ~ ., scales = "free_y") + 
  geom_ribbon(data = quarts, aes(ymin = quart1, max = quart2), 
              fill = 'grey90') +
  geom_line(size=0.3) +
  geom_point(data = mins, col = 'red') +
  geom_point(data = maxs, col = 'blue') +
  geom_text(data = mins, aes(label = Sales), vjust = -1) +
  geom_text(data = maxs, aes(label = Sales), vjust = 2.5) +
  geom_text(data = ends, aes(label = Sales), hjust = 0, nudge_x = 0.5) +
  geom_text(data = ends, aes(label = Product), hjust = 0, nudge_x = 1.0) +
  expand_limits(x = max(sales_report$Month) + 
                  (0.25 * (max(sales_report$Month) - min(sales_report$Month)))) +
  scale_x_continuous(breaks = seq(1, 12, 1)) +
  scale_y_continuous(expand = c(0.1, 0)) +
  theme_tufte(base_size = 3, base_family = "Helvetica") +
  theme(axis.title=element_blank(), axis.text.y = element_blank(), 
        axis.ticks = element_blank(), strip.text = element_blank())
```

## Static Information Dashboard Design: gt and gtExtras methods

In this section, we will create static information dashboard by using [gt](https://gt.rstudio.com/) and [gtExtras](https://jthomasmock.github.io/gtExtras/index.html) packages.

### simple bullet chart with gt and gtextras package

the following code chunk uses gt and gtextras functions to create a bullet chart

```{R}
product %>%
  gt::gt() %>%
  gt_plt_bullet(column = current, 
              target = target, 
              width = 60,
              palette = c("lightblue", 
                          "black")) %>%
  gt_theme_538()
```

### sparklines with gtExtras

the following code chunk prepares the data for sparklines visualisation with gtExtras functions

```{r}
report <- coffeechain %>%
  mutate(Year = year(Date)) %>%
  filter(Year == "2013") %>%
  mutate (Month = month(Date, 
                        label = TRUE, 
                        abbr = TRUE)) %>%
  group_by(Product, Month) %>%
  summarise(Sales = sum(Sales)) %>%
  ungroup()
```

::: callout-note
One of the requirement of gtExtras functions is that they require you to pass data.frame with list column
:::

```{R}
report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales), 
            .groups = "drop")
```

## Plotting Coffee chain sales report

The following code chunk groups the dataset by Product, then creates column 'Monthly Sales' with the list of vales from the sales column. Then gt() converts the data into a *gt table* object, and `gt_plt_sparkline()` converts the data into a line chart. `same_limit=FALSE` ensures the products do not share the same Y axis, to avoid viewing issues if some products have much higher/lower sales than others.

```{R}
report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales), 
            .groups = "drop") %>%
   gt() %>%
   gt_plt_sparkline('Monthly Sales',
                    same_limit = FALSE)
```

### Add statistics

The code chunk below first calculates summary statistics for each product

```{r}
report %>% 
  group_by(Product) %>% 
  summarise("Min" = min(Sales, na.rm = T),
            "Max" = max(Sales, na.rm = T),
            "Average" = mean(Sales, na.rm = T)
            ) %>%
  gt() %>%
  fmt_number(columns = 4,
    decimals = 2)
```

### Combine the data.frame

The code chunk below adds the summary statistics to the table

```{r}
#get sparklines 
spark <- report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales), 
            .groups = "drop")

#get summary statistics
sales <- report %>% 
  group_by(Product) %>% 
  summarise("Min" = min(Sales, na.rm = T),
            "Max" = max(Sales, na.rm = T),
            "Average" = mean(Sales, na.rm = T)
            )

#join the datasets
sales_data=left_join(sales, spark)
```

### Plot the updated table

```{r}
sales_data %>%
  gt() %>%
  fmt_number(columns = 4,
    decimals = 2) %>%
  gt_plt_sparkline('Monthly Sales',
                   same_limit = FALSE)
```

### Combining Bullet Chart and Sparklines

Bullet charts can also be added with the following code chunk

```{r}
# get bullet chart data
bullet <- coffeechain %>%
  filter(Date >= "2013-01-01") %>%
  group_by(`Product`) %>%
  summarise(`Target` = sum(`Budget Sales`),
            `Actual` = sum(`Sales`)) %>%
  ungroup() 

#join the data sets
sales_data_bullet = sales_data %>%
  left_join(bullet, by = "Product")

# display the chart
sales_data_bullet %>%
  gt() %>%
  fmt_number(columns = 4,
    decimals = 2) %>%
  gt_plt_sparkline('Monthly Sales',
                   same_limit = FALSE) %>%
  gt_plt_bullet(column = Actual, 
                target = Target, 
                width = 28,
                palette = c("lightblue", 
                          "black")) %>%
  gt_theme_538(quiet=TRUE) #apply FiveThirtyEight theme to the table output
```

## Interactive Information Dashboard Design: reactable and reactablefmtr methods

In this section, we will create interactive information dashboard using [reactable](https://glin.github.io/reactable/index.html) and [reactablefmtr](https://kcuilla.github.io/reactablefmtr/index.html) packages.

In order to build an interactive sparklines, we need to load the dataui R package by using the code chunk below.

```{R}
pacman:: p_load(dataui)
```

### Plotting interactive sparklines

the following code chunk prepares the data for interactive sparkline visualisation using reactablefmtr

```{r}
report <- report %>%
  group_by(Product) %>%
  summarize(`Monthly Sales` = list(Sales))
```

next, the sparklines are plotted as below

```{r}
reactable(
  report,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(report)
    )
  )
)
```

### Changing pagesize

Default pagesize is 10, but `defaultPageSize` can be used to change the default setting

```{r}
reactable(
  report,
  defaultPageSize = 13, #change default setting
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(report)
    )
  )
)
```

### Adding points and labels

In the code chunk below `highlight_points` argument is used to show the minimum and maximum values points and `label` argument is used to label first and last values.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        labels = c("first", "last")
        )
    )
  )
)
```

###  Adding reference line

In the code chunk below `statline` argument is used to show the mean line

::: panel-tabset
#### Mean 

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        statline = "mean"
        )
    )
  )
)
```

#### Median 

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        statline = "median"
        )
    )
  )
)
```

#### Min

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` =colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        statline = "min"
        )
    )
  )
)
```

#### `Max`

```         
{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        statline = "max"
        )
    )
  )
)
```
:::

### Adding bandline

Instead adding reference line, bandline can be added by using the `bandline` argument.

```{reactable(}
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        line_width = 1,
        bandline = "innerquartiles",
        bandline_color = "green"
        )
    )
  )
)
```

### Sparkbar 

We can also display sparkbars instead of a sparkline as seen below

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkbar(
        report,
        highlight_bars = highlight_bars(
          min = "red", max = "blue"),
        bandline = "innerquartiles",
        statline = "mean")
    )
  )
)
```
