---
title: "Visualising and Analysing Time-oriented Data"
format: html
author: "Erica Chan"
date: "02-26-2026"
date-modified: "last-modified"
number-sections: true
---

## Learning Outcome

-   plotting a calender heatmap by using ggplot2 functions,
-   plotting a cycle plot by using ggplot2 function,
-   plotting a slopegraph
-   plotting a horizon chart

## Getting started

### Import packages

```{r}
pacman::p_load(scales, viridis, lubridate, ggthemes, gridExtra, readxl, knitr, data.table, tidyverse, extrafont)
```

## Plotting Calendar Heatmap

In this section, we will:

-   plot a calender heatmap by using ggplot2 functions and extension,
-   to write function using R programming,
-   to derive specific date and time related field by using base R and lubridate packages
-   to perform data preparation task by using tidyr and dplyr packages.

### Data

#### Importing the data

First, you will use the code chunk below to import *eventlog.csv* file into R environment and called the data frame as *attacks*.

```{r}
attacks <- read_csv("data/eventlog.csv")
```

#### Examining data structure

*kable()* can be used to review the structure of the imported data.

```{r}
kable(head(attacks))
```

There are three columns, namely *timestamp*, *source_country* and *tz*.

-   *timestamp* field stores date-time values in POSIXct format.

-   *source_country* field stores the source of the attack. It is in *ISO 3166-1 alpha-2* country code.

-   *tz* field stores time zone of the source IP address.

#### Data Preparation

*Deriving weekday and hour_of_day fields*

The following code chunk creates a function to derive the desired feels

```{r}
make_hr_wkday <- function(ts, sc, tz) {
  real_times <- ymd_hms(ts, 
                        tz = tz[1], 
                        quiet = TRUE)
  dt <- data.table(source_country = sc,
                   wkday = weekdays(real_times),
                   hour = hour(real_times))
  return(dt)
  }
```

::: callout-note
-   [`ymd_hms()`](https://lubridate.tidyverse.org/reference/ymd_hms.html) and [`hour()`](https://lubridate.tidyverse.org/reference/hour.html) are from [**lubridate**](https://lubridate.tidyverse.org/) package, and
-   [`weekdays()`](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/weekdays) is a **base** R function.
:::

*Deriving the attacks tibble data frame*

```{r}
wkday_levels <- c('Saturday', 'Friday', 
                  'Thursday', 'Wednesday', 
                  'Tuesday', 'Monday', 
                  'Sunday')

attacks <- attacks %>%
  group_by(tz) %>%
  do(make_hr_wkday(.$timestamp, 
                   .$source_country, 
                   .$tz)) %>% 
  ungroup() %>% 
  mutate(wkday = factor(
    wkday, levels = wkday_levels),
    hour  = factor(
      hour, levels = 0:23))
```

::: callout-note
Beside extracting the necessary data into *attacks* data frame, `mutate()` of **dplyr** package is used to convert *wkday* and *hour* fields into **factor** so they’ll be ordered when plotting
:::

```{r}
kable(head(attacks))
```

#### Building Calendar Heatmaps

```{r}
#| fig-width: 6
grouped <- attacks %>% 
  count(wkday, hour) %>% 
  ungroup() %>%
  na.omit()

ggplot(grouped, 
       aes(hour, 
           wkday, 
           fill = n)) + 
geom_tile(color = "white", 
          linewidth = 0.1) + 
theme_tufte(base_family = "Times New Roman") + 
coord_equal() +
scale_fill_gradient(name = "# of attacks",
                    low = "sky blue", 
                    high = "dark blue") +
labs(x = NULL, 
     y = NULL, 
     title = "Attacks by weekday and time of day") +
theme(axis.ticks = element_blank(),
      plot.title = element_text(hjust = 0.5),
      legend.title = element_text(size = 8),
      legend.text = element_text(size = 6) )
```

::: callout-info
-   a tibble data table called *grouped* is derived by aggregating the attack by *wkday* and *hour* fields.
-   a new field called *n* is derived by using `group_by()` and `count()` functions.
-   `na.omit()` is used to exclude missing value.
-   `geom_tile()` is used to plot tiles (grids) at each x and y position. `color` and `size` arguments are used to specify the border color and line size of the tiles.
-   [`theme_tufte()`](https://jrnold.github.io/ggthemes/reference/theme_tufte.html) of [**ggthemes**](https://jrnold.github.io/ggthemes/reference/index.html) package is used to remove unnecessary chart junk. To learn which visual components of default ggplot2 have been excluded, you are encouraged to comment out this line to examine the default plot.
-   `coord_equal()` is used to ensure the plot will have an aspect ratio of 1:1.
-   `scale_fill_gradient()` function is used to creates a two colour gradient (low-high).
:::

#### Building Multiple Calendar Heatmaps

*Step 1: Deriving attack by country object*

In order to identify the top 4 countries with the highest number of attacks, we have to perform the following:

-   count the number of attacks by country,
-   calculate the percent of attackes by country, and
-   save the results in a tibble data frame.

```{r}
attacks_by_country <- count(
  attacks, source_country) %>%
  mutate(percent = percent(n/sum(n))) %>%
  arrange(desc(n))
```

*Step 2: Preparing the tidy data frame*

In this step, we will extract the attack records of the top 4 countries from *attacks* data frame and save the data in a new tibble data frame.

```{r}
top4 <- attacks_by_country$source_country[1:4]
top4_attacks <- attacks %>%
  filter(source_country %in% top4) %>%
  count(source_country, wkday, hour) %>%
  ungroup() %>%
  mutate(source_country = factor(
    source_country, levels = top4)) %>%
  na.omit()
```

*Step 3: Plotting the Multiple Calender Heatmap by using ggplot2 package.*

```{r}
#| fig-width: 8
ggplot(top4_attacks, 
       aes(hour, 
           wkday, 
           fill = n)) + 
  geom_tile(color = "white", 
          size = 0.1) + 
  theme_tufte(base_family = "Times New Roman") + 
  coord_equal() +
  scale_fill_gradient(name = "# of attacks",
                    low = "sky blue", 
                    high = "dark blue") +
  facet_wrap(~source_country, ncol = 2) +
  labs(x = NULL, y = NULL, 
     title = "Attacks on top 4 countries by weekday and time of day") +
  theme(axis.ticks = element_blank(),
        axis.text.x = element_text(size = 7),
        plot.title = element_text(hjust = 0.5),
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 6) )

```

## Plotting Cycle Plot

In this section, we will plot a cycle plot showing the time-series patterns and trend of visitor arrivals from Vietnam programmatically by using ggplot2 functions.

### Data Import

For the purpose of this hands-on exercise, *arrivals_by_air.xlsx* will be used.

The code chunk below imports *arrivals_by_air.xlsx* by using `read_excel()` of **readxl** package and save it as a tibble data frame called *air*.

```{r}
air <- read_excel("data/arrivals_by_air.xlsx")
```

### Deriving month and year fields

Next, two new fields called *month* and *year* are derived from *Month-Year* field.

```{r}
air$month <- factor(month(air$`Month-Year`), 
                    levels=1:12, 
                    labels=month.abb, 
                    ordered=TRUE) 
air$year <- year(ymd(air$`Month-Year`))
```

### Extracting the target country

Next, the code chunk below is use to extract data for the target country (i.e. Vietnam)

```{r}
Vietnam <- air %>% 
  select(`Vietnam`, 
         month, 
         year) %>%
  filter(year >= 2010)
```

### Computing year average arrivals by month

The code chunk below uses `group_by()` and `summarise()` of **dplyr** to compute year average arrivals by month.

```{r}
hline.data <- Vietnam %>% 
  group_by(month) %>%
  summarise(avgvalue = mean(`Vietnam`))
```

### Plotting the cycle plot

The code chunk below is used to plot the cycle plot as shown in Slide 12/23.

```{r}
#| fig-width: 16
#| fig-height: 6
ggplot() + 
  geom_line(data=Vietnam,
            aes(x=year, 
                y=`Vietnam`, 
                group=month), 
            colour="black") +
  geom_hline(aes(yintercept=avgvalue), 
             data=hline.data, 
             linetype=6, 
             colour="red", 
             size=0.5) + 
  facet_grid(~month) +
  labs(title = "Visitor arrivals from Vietnam by air, Jan 2010-Dec 2019") +
  xlab("") +
  ylab("No. of Visitors") +
  theme_gray(base_family = "Times New Roman") 
```

## Plotting Slopegraph

In this section we will plot a slopegraph by using R.

### R packages

```{r}
pacman::p_load(CGPfunctions)
```

### Data Import

The code chunk below imports the dataset into R

```{r}
rice <- read_csv("data/rice.csv")
```

### Plotting the slopegraph

The code chunk below plots a basic slopegraph

```{R}
rice %>% 
  mutate(Year = factor(Year)) %>%
  filter(Year %in% c(1961, 1980)) %>%
  newggslopegraph(Year, Yield, Country,
                Title = "Rice Yield of Top 11 Asian Counties",
                SubTitle = "1961-1980",
                Caption="Rice yield in Indonesia increased dramatically from 1961 to 1980, raising it from 7th to 4th in Asia")
```
